Modelling BCG vaccination in the UK

### 2019-08-05 rmds/10-model-results.Rmd
H model fitting and results comments

In the previous chapter I outlined a model fitting pipeline and discussed the results from using it on the model developed in Chapter \@ref(model-development). Whilst this fitted model may be used to explore the epidemiology of tuberculosis (TB) in the early 2000's it does not - as currently stands - explore the impact of the 2005 change in BCG vaccination policy (Chapter \@ref(background)).

@@

* Universal BCG vaccination of those at school-age continued with the same coverage as previously.
* Universal BCG vaccination of those at school-age was phased out in 2005 and replaced with universal BCG vaccination of neonates with the same coverage levels as assumed for the BCG schools scheme.
* Universal BCG vaccination of those at school-age was phased out in 2005.

Universal vaccination of neonates was used as a proxy for targeted vaccination of high risk neonates. This was necessary because the high risk population was not modelled in the model developed in Chapter \@ref(model-development) due to the lack of data on which to base key assumptions. No vaccination was used as a baseline in order to explore the absolute impact of vaccination. Vaccination coverage was assumed to be constant across all scenarios as there was little data on which to base between assumption variation. Regardless of the scenario considered it was assumed that school-age vaccination was in place from 1953 through to 2004.

@@

Estimated age-stratified, and aggregated, TB incidence, and mortality were compared both visually and numerically from 2005 through to 2040 for each scenario. Multiple time horizons were evaluated as initially the impact of any policy change may be masked by the large reservoir of vaccinated individuals in the population and because of the impact of the assumed decrease in non-UK born cases over time.

@@

As expected the neonatal vaccination resulted in a rapid decline in TB incidence in 0-4 year olds and a smaller but still large reduction in 5-9 year olds (Figure \@ref(fig:10-age-scenarios-fig)). There was a slight reduction in 10-15 year olds. School-age vaccination resulted in the lower incidence rates in all adult populations when compared to any other scenario, except in adults over 45 where all scenarios were comparable. Neonatal vaccination resulted in a slight decrease in TB incidence rates when compared to no vaccination in young adults.

@@

```{r 10-table-deaths-preds}
sum_deaths_scenario %>% filter(Year %% 5 == 0) %>% 
  set_names(., nm = names(.) %>% str_replace_all("BCG", "BCG (95% CI)")) %>% 
  select(-Observation) %>% 
kable(
  caption.short = "Forecast overall TB mortality for each scenario evaluated from 2005 to 2040.", 
  caption = "Forecast overall TB mortality for each scenario evaluated from 2005 to 2040. For brevity only 5 year intervals are shown.(95\\% CI): 95\\% credible interval estimated using the 2.5\\% quantile and the 97.5\\% quantile.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("HOLD_position"))
```

@@

In this chapter I outlined 3 vaccination scenarios to explore: continuing school-age BCG vaccination; universal neonatal vaccination; and no further vaccination. For each scenario I forecast TB incidence and TB mortality from 2005 through to 2040 assuming that non-UK born incidence rates would follow the same age stratified trends observed between 2010 and 2015. Although the results presented here are only preliminary, due to the quality of the model fit, it appears that continuing school-age BCG vaccination would have resulted in slightly reduced TB incidence across all time-points considered compared to any other scenario. Neonatal BCG vaccination resulted in significantly reduced TB incidence and mortality in young children but had little impact later in life, with a comparable effect to no vaccination. School-age vaccination had no impact on young children but did reduce TB incidence and mortality for all adults up to 45 years old. Beyond 45 years old no scenario impacted TB incidence or mortality.

The model developed in Chapter \@ref(model-development) was motivated by existing theory and robustly parameterised to the available data. It represents the only (known) open-source model of TB transmission and BCG vaccination. However the model fitting pipeline developed in Chapter \@ref(model-fitting) did not produce a good fit to the observed data. Ad-hoc model calibration (as discussed in Chapter \@ref(model-fitting)) failed to significantly improve on this fit. This means that the findings presented in this chapter can be considered as indicative only. However, these findings still represent the only modelling study of TB dynamics after a large scale change in BCG vaccination policy. The lack of data to support modelling the high-risk TB population population meant that targeted vaccination of high-risk neonates could not be considered as a scenario. This means that the results presented in this chapter do not contain the vaccination policy that is currently in place and so findings from the model cannot be directly compared to observed incidence data. However, only considering scenarios that alter the age of those vaccinated, rather than both the age of those vaccinated and the targeting of the vaccine, make understanding the impact of changes in vaccination policy easier to determine. The forecasts presented in this chapter are highly sensitive to the forecasted number of non-UK born TB cases. Whilst the regression method outlined in this chapter extrapolates based on current age-stratified trends it may be the case that this extrapolation breaks down over the long - or short - term. To a lesser extent the forecasts presented here are sensitive to the projected number of births and mortality rates. This is particularly the case for incidence rates in neonates and in older adults.

To my knowledge, there are no other dynamic modelling studies evaluating the use of the BCG vaccine in low burden settings that include a comparable level of detail and that are robustly parameterised based on the latest evidence. See Chapter \@ref(model-development) for a discussion of the strengths and weaknesses of the model used here in comparison with others used in the literature. However, Harris et al. recently reviewed mathematical models that explored the epidemiological impacts of future TB vaccines.[@Harris2016] They found that vaccines targeted at all-ages or at adolescents/adults were more effective at eradicating TB than neonatal programmes when vaccine effectiveness was not assumed to degrade with age. These findings agree with those presented in this Chapter, with fewer overall cases observed when vaccination continued in those at school-age, compared to neonatal vaccination. However vaccination in neonates did lead to a decrease in incidence in children both over the long - and short term - in comparison to vaccination at school-age.

@@

The results presented here indicate that changing from a school-age BCG vaccination programme to a neonatal BCG programme lead to an overall increase in incidence rates, with increases concentrated in the young adults, and to a lesser degree, in older adults. Neonatal vaccination led to a decrease in incidence rates in children both in the short - and long term. This indicates that direct vaccination provides the best protection for children rather than indirect protection via reduced transmission. This finding is likely to be dependent on the degree of background transmission and so further modelling studies are needed in diverse settings before conclusions can be generalised. No vaccination was shown to lead to increased incidence in all age-groups when compared to school-age vaccination and in children only when compared to neonatal vaccination. The impact of any vaccination programme on older adults was small. These results are preliminary in nature as the model on which they are based fitted poorly to the observed data.

>>>>

In the previous chapter I outlined a model fitting pipeline and discussed the results from using it on the model developed in Chapter \@ref(model-development). Whilst this fitted model may be used to explore the epidemiology of tuberculosis (TB) in the early 2000's it does not - as currently stands - explore the impact of the 2005 change in BCG vaccination policy (Chapter \@ref(background)). Models are useful in this context as dynamic model forecasts can be derived for multiple scenarios that may only exist on paper and so have little to no data to support them. These forecasts may then be used by policy makers as indicators of the likely impact of these scenarios.

@@

* Universal BCG vaccination of those at school-age continued with the same coverage as previously.
* Universal BCG vaccination of those at school-age was phased out in 2005 and replaced with universal BCG vaccination of neonates with the same coverage levels as assumed for the BCG schools scheme.
* Universal BCG vaccination of those at school-age was phased out in 2005 (i.e no vaccination post 2004).

The BCG policy change in 2005 was from universal school-age BCG vaccination to targeted vaccination of high risk neonates. However, here universal vaccination of neonates is used as a proxy for targeted vaccination of high risk neonates. This was necessary because the high risk population was not modelled in the model developed in Chapter \@ref(model-development) due to the lack of data on which to base key assumptions. No vaccination was used as a baseline in order to explore the absolute impact of vaccination. Vaccination coverage was assumed to be constant across all scenarios as there was little data on which to base between assumption variation. Regardless of the scenario considered it was assumed that school-age vaccination was in place from 1953 through to 2004.

@@

Estimated age-stratified, and aggregated, TB incidence, and mortality were compared both visually and numerically from 2005 through to 2040 for each scenario. Multiple time horizons were evaluated across this timespan as initially the impact of any policy change may be masked by the large reservoir of vaccinated individuals in the population and because of the impact of the assumed decrease in non-UK born cases over time.

@@

As expected the neonatal vaccination resulted in a rapid decline in TB incidence in 0-4 year olds and a smaller but still large reduction in 5-9 year olds (Figure \@ref(fig:10-age-scenarios-fig)). There was a slight reduction in 10-15 year olds. School-age vaccination resulted in the lower incidence in all adult populations when compared to any other scenario, except in adults over 45 where all scenarios were comparable. Neonatal vaccination resulted in a slight decrease in TB incidence rates when compared to no vaccination in young adults.

@@

```{r 10-table-deaths-preds}
sum_deaths_scenario %>% filter(Year %% 5 == 0) %>% 
  set_names(., nm = names(.) %>% str_replace_all("BCG", "BCG (95% CI)")) %>% 
  select(-Observation) %>% 
kable(
  caption.short = "Forecast overall TB mortality for each scenario evaluated from 2005 to 2040.", 
  caption = "Forecast overall TB mortality for each scenario evaluated from 2005 to 2040. For brevity only 5 year intervals are shown.(95\\% CI): 95\\% credible interval estimated using the 2.5\\% quantile and the 97.5\\% quantile.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("hold_position"))
```

@@

In this chapter I outlined 3 vaccination scenarios to explore: continuing school-age BCG vaccination; universal neonatal vaccination; and no further vaccination. For each scenario I forecast TB incidence and TB mortality from 2005 through to 2040 assuming that non-UK born incidence rates would follow the same age stratified trends observed between 2010 and 2015. Although the results presented here are only preliminary, due to the quality of the model fit, it appears that continuing school-age BCG vaccination would have resulted in slightly reduced TB incidence across all time-points considered compared to any other scenario. Neonatal BCG vaccination resulted in significantly reduced TB incidence and mortality in young children but had little impact later in life with a comparable effect to no vaccination. School-age vaccination had no impact on young children but did reduce TB incidence and mortality for all adults up to 45 years old. Beyond 45 years old no scenario impacted TB incidence or mortality.

The model developed in Chapter \@ref(model-development) was motivated by existing theory and robustly parameterised to the available data. It represents the only (known) open-source model of TB transmission and BCG vaccination. However, the model fitting pipeline developed in Chapter \@ref(model-fitting) did not produce a good fit to the observed data. Ad-hoc model calibration (as discussed in Chapter \@ref(model-fitting)) failed to significantly improve on this fit. This means that the findings presented in this chapter can be considered as indicative only. However, these findings still represent the only modelling study of TB dynamics after a large scale change in BCG vaccination policy. The lack of data to support modelling the high-risk TB population population meant that targeted vaccination of high-risk neonates could not be considered as a scenario. This means that the results presented in this chapter do not contain the vaccination policy that is currently in place and so findings from the model cannot be directly compared to observed incidence data. However, only considering scenarios that alter the age of those vaccinated, rather than both the age of those vaccinated and the targeting of the vaccine, make understanding the impact of changes in vaccination policy easier to determine. The forecasts presented in this chapter are highly sensitive to the forecasted number of non-UK born TB cases. Whilst the regression method outlined in this chapter extrapolates based on current age-stratified trends it may be the case that this extrapolation breaks down over the long - or short - term. To a lesser extent the forecasts presented here are sensitive to the projected number of births and mortality rates. This is particularly the case for incidence rates in neonates and in older adults.

To my knowledge, there are no other dynamic modelling studies evaluating the use of the BCG vaccine in low burden settings that include a comparable level of detail and that are robustly parameterised based on the latest evidence. Harris et al. recently reviewed mathematical models that explored the epidemiological impacts of future TB vaccines.[@Harris2016] They found that vaccines targeted at all-ages or at adolescents/adults were more effective at eradicating TB than neonatal programmes when vaccine effectiveness was not assumed to degrade with age. These findings agree with those presented in this Chapter, with fewer overall cases observed when vaccination continued in those at school-age, compared to neonatal vaccination. However vaccination in neonates did lead to a decrease in incidence in children both over the long - and short term - in comparison to vaccination at school-age.

@@

The results presented here indicate that changing from a school-age BCG vaccination programme to a neonatal BCG programme lead to an overall increase in TB incidence, with increases concentrated in the young adults, and to a lesser degree, in older adults. Neonatal vaccination led to a decrease in TB incidence in children both in the short - and long term. This indicates that direct vaccination provides the best protection for children rather than indirect protection via reduced transmission. This finding is likely to be dependent on the degree of background transmission and so further modelling studies are needed in diverse settings before conclusions can be generalised. No vaccination was shown to lead to increased incidence in all age-groups when compared to school-age vaccination and in children only when compared to neonatal vaccination. The impact of any vaccination programme on older adults was small. These results are preliminary in nature as the model on which they are based fitted poorly to the observed data. However, they do indicate some of the trade-offs involved in setting BCG policy. If reducing childhood incidence was a goal of the 2005 policy change then these results indicate a clear success. On the other hand if reducing overall TB incidence was the goal then stopping shool-age vaccination has not been a success.
